---
  title: "Bivariate Survival Outcome Analysis Using Penalized Partial Likelihood (BivPPL)"
  author: "[Lili Wang](mailto:lilywang@umich.edu)"
  date: "`r Sys.Date()`"
  output: github_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                     comment = "#>",
                     fig.width=12, fig.height=8,
                     fig.path = "man/figures/README-")

```

## Purpose

This R package is to analyze bivariate survival outcomes. Here specifically, we implement this method to analyze alternating recurrent events using a bivariate correlated frailty model. Both the regression parameters and the variance-covariance matrix will be estimated. This R package allows data to have many clusters. The estimation procedure is similar to the traditional PPL method as established in [coxme](https://cran.r-project.org/web/packages/coxme/index.html), while we do not require knowing the correlation direction between the correlated two events. 

This R package will be improved and upgraded in the near future. 


## Install the package

To install the R package from Github, you will need to install another R package "devtools". Please uncomment the codes to install them. 

```{r installation, results = "hide"}
# install.packages("devtools")
# library(devtools)
# install_github("lilywang1988/BivPPL")
library(BivPPL)
```



## Vignettes

Generate alternating recurrent event data using `gen.data`, `beta1` and `beta2` are regression parameters for the two events,  `theta` is defining the 3 different entries of the variance-covariance matrix of the two correlated frailties. `N` is the number of clusters or subjects in the framework of alternating recurrent events.

```{r vignettes}
set.seed(100)
N<-100 # number of clusters
beta1 <- c(0.5,-0.3,0.5) # regression parameters for event type 1
beta2 <- c(0.8,-0.2,0.3) # regression parameters for event type 2
beta  <- c(beta1,beta2)
theta <- c(0.25,0.25,-0.125) # variance-covariance matrix for the bivariate frailty (denoted as D), it is a vector (D[1,1],D[2,2], D[1,2])
lambda01 <- 1
lambda02 <- 1
cen <- 10 # maximum censoring time
centype <- TRUE # fixed censoring time at cen

data <- gen.data(N,beta1,beta2,theta,lambda01,lambda02,c=cen,Ctype=centype)
ptm<-proc.time()
res <- BivPPL(data)
proc.time() - ptm
res$beta_hat
res$beta_ASE
res$D_hat


# fit the model assuming the independence between the two frailties
ptm<-proc.time()
res2 <- BivPPL(data,independence=T)
proc.time() - ptm
res2$beta_hat
res2$beta_ASE
res2$D_hat

# A likelihood ratio test
LRT<-2*(res$LogMargProb-res2$LogMargProb)
print(round(pchisq(abs(LRT),df=1,lower.tail = F),3))


# sparsen the Hessian matrix
ptm<-proc.time()
res3 <- BivPPL(data,huge=TRUE)
proc.time() - ptm
res3$beta_hat
res3$beta_ASE
res3$D_hat

# Compare with coxme
# install.packages("coxme")
library(coxme)
data_coxme <- tocoxme(data) # assume that we do not know the two events are negatively associated and we transform the data to be positively associated
ptm<-proc.time()
res_coxme <- coxme(Surv(data_coxme$time,data_coxme$delta)~data_coxme$Z1+data_coxme$Z2+(1|data_coxme$b0)+(1|data_coxme$b1)+(1|data_coxme$b2)+strata(data_coxme$joint)) # disable the Hessian matrix sparsening and likelihood refining
proc.time() - ptm
res_coxme$coefficients 
sqrt(diag(vcov(res_coxme)))
assemble(as.vector(unlist(res_coxme$vcoef)))

data_coxme_n <- tocoxme_n(data) # assume that we know the two events are negatively associated 
ptm<-proc.time()
res_coxme_n<- coxme(Surv(data_coxme_n$time,data_coxme_n$delta)~data_coxme_n$Z1+data_coxme_n$Z2+(data_coxme_n$Z0|data_coxme_n$b0)+(1|data_coxme_n$b1)+(1|data_coxme_n$b2)+strata(data_coxme_n$joint))
proc.time() - ptm
res_coxme_n$coefficients 
sqrt(diag(vcov(res_coxme_n)))
assemble_n(as.vector(unlist(res_coxme_n$vcoef)))
```
